---
title: "Tidy Tuesday at R-Ladies Melbourne Inc: data wrangling and visualisation of the measles vaccination rate in the US"
author: Momeneh (Sepideh) Foroutan
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---


```{r, include = F}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


Measles is a very contagious respiratory infection, and [this tidy tuesday data](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-25/readme.md) has vaccination rates for 46,412 schools in 32 states in the US, inluding overall vaccination rate, Measles, Mumps, and Rubella (MMR) vaccination rate, as well percent of student who were exempted form the vaccination due to different reasons. 

First we read in the data, and load the `tidyverse` library.
```{r, warning = F,  message = F}
measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-25/measles.csv')

library(tidyverse)
```


# Explore the data
Looking at the dimension, head, structure and summary of the data, gives us some insight about the datasets. Bases on these, we realise that there are -1 in the mmr and overall columns, which we may want to take this into account when visualising the data. You may also have noticed that one of the exemption columns is logical instead of numeric.
This also help to see which columns have how many NAs.
```{r}
dim(measles)
head(measles)
str(measles)
summary(measles)
```

We can also look at the tables of different columns. Looking at this, we see that the year column has some grouping issues, that you may want to change.
```{r}
selCols <- c("state", "year", "type")
apply(measles[, selCols], 2, table )
```

Now, to do some visualisations, lets calculate the average of the vaccination rates and different expemtion columns. 

```{r}
mstat <- measles %>%
  group_by(state) %>%
  mutate(ave_xmed = mean(xmed, na.rm = T),
         ave_xper = mean(xper, na.rm = T),
         ave_xrel = mean(xrel, na.rm = T),
         ave_enroll = mean(enroll, na.rm = T),
         ave_mmr = mean(mmr, na.rm = T),
         ave_overall = mean(overall, na.rm = T)) %>%
  data.frame() 
```

# Visualisation
## Barplot
In order to make barplots for the vaccination rate columns, we make sure that we filter out -1 values from that column. As we have calculated the mean vaccination rate values per state, we remove duplicated rows too.
Note that we can order the barplots, simply by using the `reorder()` function.
```{r}
mstat %>%  
  filter(! duplicated(state) & ave_overall != -1) %>% 
ggplot(., aes(x = reorder(state, ave_overall), y = ave_overall)) + 
  geom_bar(stat = "identity", fill = "gray20") +
  ggtitle("Average overall vaccination across states") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mstat %>%  
  filter(! duplicated(state) & ave_mmr != -1) %>% 
ggplot(., aes(x = reorder(state, ave_mmr), y = ave_mmr)) + 
  geom_bar(stat = "identity", fill = "gray60") +
  ggtitle("Average MMR vaccination across states") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Now, if we want to know what percent exemption of different reasons we have in each state, we should first change the structure of the data to be in the long format, so that all teh info for different exemptions go into one column. For this, we use `pivot_longer()` function. The output of this can be directly used in the `ggplot()` function. Note that we also filter for NaN valuse in the new column "Average_values".
```{r}
mstat %>%
  filter(!duplicated(state)) %>%
  pivot_longer(.,
               cols = ave_xmed:ave_xrel,
               names_to = "Exemption",
               values_to = "Average_values") %>%
  filter(!is.nan(Average_values)) %>%
  ggplot(., aes(
    x = reorder(state, ave_mmr),
    y = Average_values,
    fill = Exemption
  )) +
  geom_bar(stat = "identity",
           position = "dodge",
           width = 0.8) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Scatterplot
The other way of looking at the data is to examine associations. For example, there seems to be a positive correlation between overall vaccination rate and mmr, however, there are schools in some states, such as Colorado that have high mmr but generally lower overall vaccination rate.
```{r}
mstat %>%
  filter(overall != -1 & mmr != -1) %>%
  ggplot(., aes(x = mmr, y = overall, color = state)) +
  geom_point(alpha = 0.6) +
  theme_bw()
```

## Map 
We can also examine this data in form of a map. There are several ways of doing this but here I am giving two simple examples to get started. These include using the `coord_quickmap()` function from ggplot2 and `plot_usmap()` function from usmap package.

### coord_quickmap
Here, we plot all lng and lat, and then because there seem to be some outliers in these data, we focus on only those lng and lat that properly covers the US coordinates. Then we show how to simply color that based on state.
```{r}
ggplot(mstat, aes(lng, lat)) + 
  geom_point(size = .25, show.legend = FALSE) +
  coord_quickmap()

mstat %>% 
  filter(lng < -60 ) %>% 
ggplot(., aes(lng, lat)) + 
  geom_point(size = .25, show.legend = FALSE) +
  coord_quickmap()


mstat %>% 
  filter(lng < -60 ) %>% 
ggplot(., aes(lng, lat, color = state)) + 
  geom_point(size = .25, show.legend = FALSE) +
  coord_quickmap()
```

We can also colour the points based on some continuous values, such as mmr or overall. To have a better contrast in the colour, I decided to keep -1 values, however, you can remove it and see what your plot looks like.
```{r}
mstat %>%
  filter(lng < -60) %>%
  # filter(mmr != -1) %>%
  ggplot(., aes(lng, lat, color = mmr)) +
  geom_point(size = .25, show.legend = TRUE) +
  scale_color_viridis_c(name = "MMR", label = scales::comma) +
  coord_quickmap() +
  theme(legend.position = "right") +
  theme_dark()

mstat %>%
  filter(lng < -60) %>%
  # filter(overall != -1) %>%
  ggplot(., aes(lng, lat, color = overall)) +
  geom_point(size = .25, show.legend = TRUE) +
  scale_color_viridis_c(name = "Overall", label = scales::comma) +
  coord_quickmap() +
  theme(legend.position = "right") +
  theme_dark()

```

### Make it interactive
Now, we subset the data to those that do not have -1 in overall column and color based on the mmr values. We can zoom on different states and make the plots inteactive using the plotly package and `ggplotly()` function. Uncomment `plotly::ggplotly(p, tiptools = "text")` in the below code in your computer to see the interactivity.
```{r}
p <- mstat %>%
  filter(lng < -110) %>%
  filter(overall != -1) %>%
  ggplot(., aes(lng, lat, color = mmr, text = name)) +
  geom_point(size = .25, show.legend = TRUE) +
  scale_color_viridis_c(name = "MMR", label = scales::comma) +
  coord_quickmap() +
  theme(legend.position = "right") +
  theme_dark()

p
# plotly::ggplotly(p, tiptools = "text")
```

### usmap
As I mentioned above, we can also use usmap package to generate maps for the US. To do this, I have been looking at the post [here](https://cran.r-project.org/web/packages/usmap/vignettes/mapping.html); but there are obviously other relevant posts, such as [this](https://remiller1450.github.io/s230s19/Intro_maps.html) that shows how to generate maps using ggplot2, for example using `geom_polygon()` function.

Note that as we do not have state abbreviations, we use state.name and state.abb from the usmap, merge these to our data and use `plot_usmap()` to visualise it.

```{r}
library(usmap)

mstatUniqueState <- mstat[! duplicated(mstat$state), ]
abbr_state <- tibble(state = state.name, abb = state.abb)
  
mstatUniqueState %>% 
  filter(ave_mmr != -1) %>% 
  left_join(abbr_state) %>% 
plot_usmap(
  data = .,
  region = "state",
  values = "ave_mmr",
  color = "purple"
) +
  scale_fill_viridis_c(name = "MMR", label = scales::comma) +
  theme(legend.position = "right") +
  theme(panel.background = element_rect(color = "white", fill = "gray10"))

mstatUniqueState %>% 
  filter(ave_overall != -1) %>% 
  left_join(abbr_state) %>% 
plot_usmap(
  data = .,
  region = "state",
  values = "ave_overall",
  color = "purple"
) +
  scale_fill_viridis_c(name = "Overall", label = scales::comma) +
  theme(legend.position = "right") +
  theme(panel.background = element_rect(color = "white", fill = "gray10"))

```

# Session info
```{r}
sessionInfo()
```


<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 598px;"></div>
